Poker Chips, Earthquakes, and Computation

### 2014-10-11 finaldraft.tex
Unbotched It
Edited First 3 sections

\documentclass[runningheads,a4paper]{llncs}

@@

First, we formalize the definition of a board. The board is an infinite grid graph with a nonnegative number of chips placed at each node
\begin{center}
\includegraphics[scale=0.5]{grid1.png}
\end{center}

\begin{center}
\includegraphics[scale=0.8]{cartesiangrid.png}
\end{center}
We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$. We will define the board as a function from the Cartesian grid to the nonnegative integers.

@@

This function maps node labels given as integer coordinates to chip stack sizes. For example, FIGUREREFERENCE if a board $b$ has 8 chips at the node with coordinate $(3,1)$, then $b(3,1) = 8$.

@@

In this section we show that any finite board will reach a stable state after finitely many earthquakes.
We do so by showing that if a earthquake changes the state of a board then no sequence of earthquakes can transform the output board back into the original board. 
We will then give a bound for the maximum number of configurations that earthquakes can transform a given board into. 
Together these show that all finite boards stabilize. 
\begin{theorem}
\label{finitestability}
If $b \in \mathcal{B}_F$, there exists $N> 0$ such that after $N$ earthquakes the resultant board will be stable, or for all $m > 0$ and all $s \in \mathbb{Z} \times \mathbb{Z}$, $E^{N+m}[b,s] = E^N[b,s]$.
\end{theorem}

@@

\begin{lemma}
Suppose $b \in \mathcal{B}_F$. If $E[b, s] \neq b$, then $E^n[b, \{s_n\}] \neq b$ for all $n > 0$ and all sequences $\{s_n\}$.
\end{lemma}

\begin{proof}
For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as $S: \mathcal{B}_F \rightarrow \mathbb{Z}_{\geq 0}$ 
\begin{equation*}
S(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
\end{equation*}
We can think of $S(b)$ as the sum of the number of chips on each square weighted by the "taxi-cab" distance from the origin squared. This is well-defined since the total number of chips on the board is finite. We claim 
\begin{equation*}
E[b, s] \neq b \Rightarrow S(E[b, s]) > S(b)
\end{equation*}
That is if an earthquake changes the board, $S$ increases. This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.

We evaluate $S(E[b, s]) - S(b)$, recalling that $E[b, s](i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
\begin{align*}
S(E[b, s]) - S(b) &=\sum_{i,j} E[b, s](i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
&=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
\end{align*}
For most two of the neighbors of $(s_1, s_2)$ 
\begin{equation*}
(|i|+|j|)= (|s_1|+|s_2|-1)^2,
\end{equation*}
and for at least two of the neighbors 
\begin{equation*}
(|i|+|j|)=(|s_1|+|s_2|+1)^2,
\end{equation*}
 and $(|s_1|+|s_2|+1)^2 > (|s_1|+|s_2|-1)^2$. Therefore 
\begin{align*}
&-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2 \\
& \geq 4(|s_1| + |s_2|)^2 + 2 (|s_1| + |s_2|+1)^2 + 2 (|s_1| + |s_2|-1)^2 = 4 >0.
\end{align*}
\end{proof}

Now that we have shown that an earthquake can never transform a finite board into a board it was previously we will give a bound on the number of different boards that any number of earthquakes can transform a given finite board into.
\begin{lemma}
\label{finiteextension}
Suppose $\sum_{i,j} b(i,j) = C$, and $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then   $E^n[b, \{s_k\}]$ is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$for all $n$ and for all sequences $\{s_k\}$ .
\end{lemma}

\begin{proof}
Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E[b, s](i,j) \geq 1$. This is clear since any earthquake will move either bring things into the column $i$ or if it moves some chips out, then there are two chips that remain in that column.

The same is true for each row. This means that if a chip ever goes to a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
\end{proof}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Recall that $b \in \mathcal{B}_F$, so $\sum_{i,j} b(i,j) = C$. Because $b$ is a finite board we know the locations with a nonzero number of chips are bounded.  Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then by Lemma~\ref{finiteextension}, we have that $E^n[b, \{s_k \}]$ is nonzero in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$, for all $n>0$ and all sequences $\{ s_k \}$ of length $n$.
If we set 
\begin{equation*}
D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
\end{equation*}
then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,. We know that the state of the board cannot cycle, so there must be an $0 \leq N \leq \binom{D+C-1}{D-1}$ and some sequence $\{s_k\}$, which is arbitrary after the first $N$ terms, such that $E^N[ b, \{s_k\}] = E^{N+m}[b, \{s_k\}]$. 
\end{proof}

\subsection{Acyclicity of Infinite Boards}
\label{Acyclicity of Infinite Boards}
\emph{written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\

We can in fact show that infinite boards cannot cycle under earthquakes.

\begin{corollary}
All boards are can never return to their same state.
\end{corollary}

\begin{proof}
We have shown that finite boards never return to the same state. Suppose that an infinite board $b$ reurns to the same state, then it does so after $N$ earthquakes at locations $\{ s_k \}$. However, we can bound the size of the locations in $\{ s_k \}$ and so this infinite board acts in the same way that a finite board on this bound acts. Therefore, if the infinite board cycles, then the finite board cycles as well. 
\end{proof}

>>>>

documentclass[runningheads,a4paper]{llncs}

@@

First, we formalize the definition of a board. The board is an infinite grid graph with a nonnegative number of chips placed at each node
\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{grid1.png}
\end{center}
\caption{A finite component of the grid}
\end{figure}

\begin{figure}
\label{cartesiangrid}
\begin{center}
\includegraphics[scale=0.8]{cartesiangrid.png}
\end{center}
\caption{A finite component of the grid with coordinates}
\end{figure}
We can coordinatize the grid graph by labeling it like the standard Cartesian grid, where we fix one node to be the origin $(0,0)$, and we index all other nodes by integer coordinates $(i,j)$. We will define the board as a function from the Cartesian grid to the nonnegative integers.

@@

This function maps node labels given as integer coordinates to chip stack sizes. For example, in Figure \ref{cartesiangrid} the board $b$ has 8 chips at the node with coordinate $(3,1)$, so we write $b(3,1) = 8$.

@@

In this section we show that any finite board will reach a stable state after finitely many earthquakes.
We do so by showing that if a earthquake changes the state of a board then no sequence of earthquakes can transform the output board back into the original board. 
We will then give a bound for the maximum number of configurations that earthquakes can transform a given board into. 
Together these show that all finite boards stabilize. 
\begin{theorem}
\label{finitestability}
If a board $b$ is finite then there exists $N>0$ such that after applying a $N$ earthquakes the resultant board will be stable.
\end{theorem}

@@

\begin{lemma}
Suppose $b$ is a finite board. If $E_{s}b \neq b$, then no application of earthquakes to $E_{s}b$ can produce $b$. 
\end{lemma}

\begin{proof}
For ease of notation we use $s=(s_1, s_2) \in \mathbb{Z} \times \mathbb{Z}$. We define our spread function as a map from finite boards to the nonnegative integers
\begin{equation*}
S(b) = \sum_{i,j} b(i,j)(|i|+|j|)^2. 
\end{equation*}
We can think of $S(b)$ as the sum of the number of chips on each square weighted by the "taxi-cab" distance from the origin squared. This is well-defined since the total number of chips on the board is finite. We claim that 
\begin{equation*}
E_{s} b \neq b \Rightarrow S(E[b, s]) > S(b)
\end{equation*}
That is if an earthquake changes the board, $S$ increases. This implies that if an earthquake changes a board, then no subsequent application of earthquakes can transform it back to its original state.

We evaluate $S(E_{s} b) - S(b)$, recalling that $E_s b(i,j)$ refers to the value of $b$ at $(i,j)$ after an earthquake effected $s$.
\begin{align*}
S(E_{s} b) - S(b) &=\sum_{i,j} E_{s} b (i,j)(|i|+|j|)^2 - \sum_{i,j} b(i,j)(|i|+|j|)^2 \\
&=-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2
\end{align*}
For at most two of the neighbors of $(s_1, s_2)$ 
\begin{equation*}
(|i|+|j|)= (|s_1|+|s_2|-1)^2,
\end{equation*}
and for at least two of the neighbors 
\begin{equation*}
(|i|+|j|)=(|s_1|+|s_2|+1)^2,
\end{equation*}
 and $(|s_1|+|s_2|+1)^2 > (|s_1|+|s_2|-1)^2$. Therefore 
\begin{align*}
&-4(|s_1| + |s_2|)^2 +\sum_{(i,j) \in n(s)} (|i| + |j|)^2 \\
& \geq 4(|s_1| + |s_2|)^2 + 2 (|s_1| + |s_2|+1)^2 + 2 (|s_1| + |s_2|-1)^2 = 4 >0.
\end{align*}
\end{proof}

Now that we have shown that an earthquake can never transform a finite board into a board it was previously we will give a bound on the number of different boards that earthquakes can transform a given finite board into.
\begin{lemma}
\label{finiteextension}
Suppose $\sum_{i,j} b(i,j) = C$, and $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then after applying any number of earthquakes the resultant board is nonzero only in the range $[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$ .
\end{lemma}

\begin{proof}
Suppose that $\sum_{j\in \mathbb{Z}} b(i,j) \geq 1$ for some $i \in \mathbb{Z}$, then $\sum_{j \in \mathbb{Z}} E_{s}b(i,j) \geq 1$. This is clear since an earthquake can bring things into column $i$ or if it moves some chips out, then there are two chips that remain in the column.

The same is true for each row. This means that if a chip  is ever in a row or column of nodes, there will be always be a chip on that row or column. This finishes the lemma. 
\end{proof}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Recall that $b$ is a finite board so $\sum_{i,j} b(i,j) = C$, furthermore this tells us that the locations with a nonzero number of chips are bounded.  Without loss of generality we can say that $b$ is nonzero only in the range $[i_{min}, i_{max}] \times [j_{min}, j_{max}]$. Then by Lemma~\ref{finiteextension}, we have that after any number of earthquakes the resultant board is nonzero only in the range $ R=[i_{min}-C, i_{max}+C] \times [j_{min}-C, j_{max}+C]$. If we set 
\begin{equation*}
D= |j_{max}-j_{min} + 2C|*|i_{max} - i_{min} + 2C|
\end{equation*}
then there are at most  $\binom{D+C-1}{D-1}$  boards that have exactly $C$ chips on them and are nonzero only in $R$,. We know that the state of the board cannot cycle and there are only a finite number of board states so eventually the board must stabilize.
\end{proof}

We can use our result for finite boards to show that earthquakes will never cause infinite boards to revisit a prior arrangement.

\begin{corollary}
Suppose $b$ is an infinite board. If $E_s b \neq b$, then no application of earthquakes to $E_s b$ can produce $b$.
\end{corollary}

\begin{proof}
We will proceed by contradiction. Assume otherwise, so there exists some sequence $\{s_1, s_2, \ldots, s_k\}$ such that after applying an earthquake to each of these locations on $E_s b$ the resultant board is $b$.
However, we can bound the size of the locations affected by $\{s, s_1, s_2, \ldots, s_k \}$, call it $R$.  This infinite board behaves the same way as a finite board with the same values on $R$ and 0 outside of $R$. But this means that earthquakes transformed a finite board back into a previous state which is a contradiction. Therefore infinite boards cannot be transformed back into a previous state.
\end{proof}
